CUBE: A Memory-lite Cosmological $N$-body Algorithm

### 2017-09-29 ms_caf/ms_caf.tex
memory section

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} uses cubic decomposition structures. The global simulation volume is decomposed into ${\tt nn}^3$ cubic sub-volumes, and each of these are assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into ${\tt nnt}^3$ cubic {\it tiles}, which is a essential unit in the calculations. Each tile is surrounded by a {\it buffer} region which is ${\tt ncb}$ coarse cells thick. The buffer is designed for two reasons: (1) computing the fine mesh force, whose cut-off length ${\tt nforce\_cutoff}\leq{\tt ncb}$, and (2) collecting all possible particles travelling from a tile's buffer region to its center, {\it physical} region. Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with ${\tt nn=2}$ and ${\tt nnt=2}$. According to this spatial decomposition, and as discussed in the last two subsections, we declare $\{\rho_c,\chi_d,v_c,\nu_d\}$, or in variable expression $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$ as follows:
\begin{eqnarray*}
	&&{\tt integer(1)\ rho\_c(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\chi{\tt )\ xp(npmax)[nn,nn,*]}\\
	&&{\tt real(4)\ vfield(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\nu{\tt )\ vp(npmax)[nn,nn,*]}
\end{eqnarray*}
where ${\tt nex}={\tt nt}+2\times{\tt ncb}$ covers the buffer region on both sides, {\tt nnt} is the tile dimemsions, and {\tt nn} is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. We denote actual number of particles in a given image {\tt nplocal}, and ${\tt npmax}>{\tt nplocal}$ is a value large enough to store buffered particles and should take into account the fact that {\tt nplocal} on some images (with densely clustered structures) could be large. The particle position and velocity arrays {\tt xp} and {\tt vp} (equivalent to $\chi_d$ and $\nu_d$ in \ref{ss.position} and \ref{ss.velocity} respectively) are required to be sorted according to the same memory layout of {\tt rho\_c}, such that $n_c$ and thus global positions of particles $x_d$ can be obtained by equation (\ref{eq.x}). $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$ provides a complete information of positions and velocities of particles, and we call it a snapshot, or checkpoint.

@@

{\tt initialize} and {\tt read\_particles} -- The subroutine {\tt initialize} creates necessary FFT plans and read in configuration files telling the program at which redshifts we need to do checkpoints, halofinds, or stop the simulation. Force kernels are also created or read in here. In {\tt read\_particles}, for each image, we read in all particles in {\it physical} regions of every tile (indicated in Figure \ref{fig.tile}), i.e., $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$. These are obtained by the initial condition generator (see Appendix A). Because physical regions of tile are {\it complete} and {\it disjoint} in space, particles at this stage are also complete and disjoint. We call it ``disjoint state''. At this stage, {\tt rho\_c}'s buffer regions of each tile are 0, and the elements of {\tt xp} and {\tt vp} beyond physical number -- {\tt xp(nplocal+1:)} and {\tt vp(nplocal+1:)} can be arbitrary and are not used.

@@

In order to keep particles ordered, for each tile, we first do $x_d=x_d+v_d\times{\tt dt}$ on all particles to obtain an updated density and velocity field on the tile, {\tt rho\_c\_new} and {\tt vfield\_new}. Then, this calculation in done on the same tile again\footnote{This repetition scales as $o(N)$ and is computational inexpensive.} to generate a new, local particle list {\tt xp\_new} and {\tt vp\_new} on the tile by Equations(\ref{eq.chi},\ref{eq.nu}). Here, the ordering of particles depends on {\tt rho\_c\_new}, and {\tt vfield\_new} is used as $v_c$ in Equation(\ref{eq.nu}). Then, another iteration is done on this tile to discard buffer regions of $\{{\tt rho\_c\_new},{\tt xp\_new},{\tt vfield\_new},{\tt vp\_new}\}$, converting buffer state to disjoint state, and it replaces corresponding part of $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$. When all tiles are iterated, the entire image is in disjoint state, and {\tt nplocal} is updated. Then all images is synchronized by {\tt sync all} and we sum up {\tt nplocal} over images to check if total number of particles {\tt npglobal} is conserved.

@@

@@

By far, {\tt vp} in physical regions are correctly updated. Remember that particle locations in the buffer regions has updated before {\tt PM} and remained unchanged. So we simply call {\tt buffer\_v} again such that the {\tt update\_x} in the next iteration will be done correctly. We call {\tt update\_vp} to convert {\tt vp} into buffered state.

@@

\subsection{Memory layout}\label{ss.memory}
The memory overhead of fine force calculation is controlled by $(1/{\tt nnt})^3$. The memory usage of coarse arrays are negligible. 
Although {\tt xp\_new} and {\tt vp\_new} arrays take extra memory, they do not appear simultaneously for multiple tiles, and compared to {\tt xp} and {\tt vp}, the extra memory overhead is by a factor of $1/{\tt nnt}^3$.
Here we summarize the memory usage in unit of bpp (byte per particle).

@@

\begin{figure}
\centering
  \includegraphics[width=0.95\linewidth]{diff_power.pdf}
 \caption{power spectrum and cross correlation.}
\label{fig.ccc}
\end{figure}

\section{Discussion and Conclusion}\label{s.discussion}
Cosmological simulation, PM method.

>>>>

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} uses cubic decomposition structures. The global simulation volume is decomposed into ${\tt nn}^3$ cubic sub-volumes with {\tt nc} coarse grids per side, and each of these are assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into ${\tt nnt}^3$ cubic {\it tiles}, with {\tt nt}={\tt nc}/{\tt nnt} coarse grids per side, which is a essential unit in the calculations. Each tile is surrounded by a {\it buffer} region which is ${\tt ncb}$ coarse cells thick. The buffer is designed for two reasons: (1) computing the fine mesh force, whose cut-off length ${\tt nforce\_cutoff}\leq{\tt ncb}$, and (2) collecting all possible particles travelling from a tile's buffer region to its center, {\it physical} region. Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with ${\tt nn=2}$ and ${\tt nnt=2}$. According to this spatial decomposition, and as discussed in the last two subsections, we declare $\{\rho_c,\chi_d,v_c,\nu_d\}$, or in variable expression $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$ as follows:
\begin{eqnarray*}
	&&{\tt integer(1)\ rho\_c(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\chi{\tt )\ xp(npmax)[nn,nn,*]}\\
	&&{\tt real(4)\ vfield(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\nu{\tt )\ vp(npmax)[nn,nn,*]}
\end{eqnarray*}
where ${\tt nex}={\tt nt}+2\times{\tt ncb}$ covers the buffer region on both sides, {\tt nnt} is the tile dimemsions, and {\tt nn} is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. We denote actual number of particles in a given image {\tt nplocal}, and ${\tt npmax}>{\tt nplocal}$ is a value (discussed in Section \ref{ss.memory}) large enough to store particles. The particle position and velocity arrays {\tt xp} and {\tt vp} (equivalent to $\chi_d$ and $\nu_d$ in \ref{ss.position} and \ref{ss.velocity} respectively) are required to be sorted according to the same memory layout of {\tt rho\_c}, such that $n_c$ and thus global positions of particles $x_d$ can be obtained by equation (\ref{eq.x}). $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$ provides a complete information of positions and velocities of particles, and we call it a snapshot, or checkpoint.

@@

{\tt initialize} and {\tt read\_particles} -- The subroutine {\tt initialize} creates necessary FFT plans and read in configuration files telling the program at which redshifts we need to do checkpoints, halofinds, or stop the simulation. Force kernels {\tt kern\_c}, {\tt kern\_f} are also created or read in here. In {\tt read\_particles}, for each image, we read in all particles in {\it physical} regions of every tile (indicated in Figure \ref{fig.tile}), i.e., $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$. These are obtained by the initial condition generator (see Appendix A). Because physical regions of tile are {\it complete} and {\it disjoint} in space, particles at this stage are also complete and disjoint. We call it ``disjoint state''. At this stage, {\tt rho\_c}'s buffer regions of each tile are 0, and the elements of {\tt xp} and {\tt vp} beyond physical number -- {\tt xp(nplocal+1:)} and {\tt vp(nplocal+1:)} can be arbitrary and are not used.

@@

In order to keep particles ordered, for each tile, we first do $x_d=x_d+v_d\times{\tt dt}$ on all particles to obtain an updated density and velocity field on the tile, {\tt rho\_c\_new} and {\tt vfield\_new}. Then, this calculation in done on the same tile again\footnote{This repetition scales as $o(N)$ and is computational inexpensive.} to generate a new, local particle list {\tt xp\_new} and {\tt vp\_new} on the tile by Equations(\ref{eq.chi},\ref{eq.nu}). Here, the ordering of particles depends on {\tt rho\_c\_new}, and {\tt vfield\_new} is used as $v_c$ in Equation(\ref{eq.nu}). Then, another iteration is done on this tile to discard buffer regions of $\{{\tt rho\_c\_new},{\tt xp\_new},{\tt vfield\_new},{\tt vp\_new}\}$, converting buffer state to disjoint state, and it replaces corresponding part of $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$. When all tiles are iterated, the entire image is in disjoint state, and {\tt nplocal} is updated. Then all images is synchronized by {\tt sync all} and we sum up {\tt nplocal} over images to check if total number of particles {\tt npglobal} is conserved. These steps are summarized in Figure \ref{fig.update_xp}.

subroutine update_xp
  do (each tile)
    do (each coarse grid)
      do (each particle)
        calculate particle's new coarse grid position
        update rho_c_new
        update vfield_new
      enddo
    enddo
    do (each coarse grid)
      do (each particle)
        calculate particle's new accurate position
        calculate particle's index
        update xp_new(index) & vp_new(index)
      enddo
    enddo
    do (each coarse grid)
      discard buffer information
      replace xp and vp for this tile
    enddo
  enddo
  sync all
  update velocity dispersion
  check nplocal & npglobal
end
\end{verbatim}
\caption{Pseudocode for subroutine {\tt update\_xp}.}
\label{fig.update_xp}
\end{figure}

@@

subroutine update_vp
  do (each tile)
    calculate fine mesh density
    calculate fine mesh force: local FFT
    update fine mesh velocity
    update maximum acceleration
    if (PP force) then
      PP calculation & velocity update
      update maximum acceleration
    endif
  enddo
  sync all
  calculate coarse mesh density
  calculate coarse mesh force: global pencil FFT
  update coarse mesh velocity
  update maximum acceleration
  update maximum velocity
end
\end{verbatim}
\caption{Pseudocode for subroutine {\tt update\_vp}.}
\label{fig.update_vp}
\end{figure}

@@

By far, {\tt vp} in physical regions are correctly updated. Remember that particle locations in the buffer regions has updated before {\tt PM} and remained unchanged. So we simply call {\tt buffer\_v} again such that the {\tt update\_x} in the next iteration will be done correctly. We call {\tt update\_vp} to convert {\tt vp} into buffered state. These steps are summarized in Figure \ref{fig.update_vp}.

@@

\subsection{Memory layout}\label{ss.memory}
Here we list the memory-consuming arrays and how they scale with different configurations of the simulation. We classify them into (1) arrays of particles, (2) coarse mesh arrays and (3) fine mesh arrays.

\subsubsection{Arrays of particles}
The arrays of particles contains {\tt xvp(6,npmax)} and a temporary {\tt xvp\_new(6,nptile)}\footnote{We use ``{\tt xvp}'' to represent $\{{\tt xp},{\tt vp}\}$, and use ``{\tt xvp\_new}'' to represent $\{{\tt xp\_new},{\tt vp\_new}\}$.}. {\tt npmax} is the number of particles {\tt xvp} can store:
\begin{equation}
	{\tt npmax}=\left\langle {\tt nplocal} \right\rangle \left( 1+\frac{2\times {\tt ncb}}{\tt nt} \right)^3(1+\epsilon_{\rm image}),
\end{equation}
where $\left\langle {\tt nplocal} \right\rangle$ is the average number of particles per image. The second term above let us store particles in buffer regions, and the third term $1+\epsilon_{\rm image}$ takes into account the inhomogeneity of {\tt nplocal} on different images. When each image models smaller physical scales, $\epsilon_{\rm image}$ should be set larger.

{\tt xvp\_new} stores temporary particles only on tiles:
\begin{equation}
	{\tt nptile}=\left\langle {\tt nplocal} \right\rangle \left( \frac{1}{\tt nnt} \right)^3(1+\epsilon_{\rm tile}),
\end{equation}
where, recall that {\tt nnt} is the number of tiles per image per dimension, and similarly $\epsilon_{\rm tile}$ controls the inhomogeneity on scale of tiles. Larger {\tt nnt} causes more inhomogeneity on smaller tiles, and $\epsilon_{\rm tile}$ can be much larger than $\epsilon_{\rm image}$, however the term ${\tt nnt}^{-3}$ decreases much faster. So, the majority memory is occupied by {\tt xvp}.

\subsubsection{Coarse mesh arrays}
On coarse mesh, {\tt rho\_c}, {\tt vfield} and force kernel {\tt kern\_c} should always be kept. They are usually configured to be 4 times coarser than fine grids and particle number density. In this case, each coarse grid, 3D {\tt vfield} takes only 12 bytes, and 64 particles will at least take 384 bytes ($n_\chi=n_\nu=1$ case). Similarly, pencil-FFT arrays, coarse force arrays etc. are memory-light and temporary. The total coarse mesh arrays take less than 10\% of the memory of {\tt xvp}.

\subsubsection{Fine mesh arrays}
On the local fine mesh, only a force kernel array needs to be kept. The fine mesh density arrays, force field arrays are temporary. Since {\tt xvp\_new}, coarse force arrays, pencil-FFT arrays are also temporary and are not used in any calculation simultaneously, they can be overlapped in memory by using equivalent statements. The total fine mesh arrays take less than 5\% of the memory of {\tt xvp}.

\subsubsection{Compare with traditional algorithms}
To show the improvement of memory usage, we give an example of TianNu simulation \citep{2017NatAs...1E.143Y} on the Tianhe-2 supercomputer, which used a traditional {\tt CUBEP3M} code. TianNu's particles number is limited by memory per computing node -- per computing node, an average of $576^3$ neutrino particles and $288^3$ CDM particles are used, and consumes about 40 GB of memory\footnote{Additional memory is used for OpenMP parallelization and for particle IDs to differentiate different particle spices.}, or about 186 bpp. A memory efficient case of {\tt CUBEP3M} by using large physical scales and at costs of speed, still uses about 40 bpp.

By using {\tt CUBE}, with parameters $n_\chi=n_\nu=1$, ${\tt nc}=384$, ${\tt nnt}=4$, ${\tt ncb}=6$, $\epsilon_{\rm image}=10\%$, $\epsilon_{\rm tile}=100\%$ and $\left\langle {\tt nplocal} \right\rangle=1536^3$, we use about 38.3 GB of memory, corresponding to 10.6 bpp. This can be done on most of the supercomputers, even laptops.

\begin{table}[]
\centering
\caption{Memory layout for a certain configuration}
\label{t.memory}
\begin{tabular}{llrrr}
\hline
& & \multicolumn{3}{c}{Memory usage}\\
\cline{3-5}
Type     & Array & /GB & /bpp & Percentage \\
\hline
Particles  & \tt xvp          & 29.9   & 8.24   & 87.4\%     \\
           & \tt xvp\_new     & \underline{1.69}   & \underline{0.465}  & \underline{4.93\%}`     \\
           & \tt PID          & 0      & 0      &    0\%     \\
           & Subtotal         & 31.5   & 8.71   & 92.4\%     \\
\hline
Coarse mesh& \tt rho\_c       & 0.296  & 0.0818 & 0.870\%    \\
           & \tt vfield       & 0.889  & 0.245  & 2.60\%     \\
           & \tt kern\_c      & 0.340  & 0.0939 & 0.996\%    \\
           & \tt force\_c     & (0.690)  & (0.190)  & (2.02\%)    \\
           & \tt rho\_c\_new  & 0.0140 & 0.00388  & 0.0411\% \\
           & Pencil-FFT       & (0.454)  & (0.125) & (1.33\%)      \\
           & Subtotal         & 1.55   & 0.429 & 4.55\%     \\
\hline
Fine mesh  & \tt kern\_f      & 1.06   & 0.292 & 3.09\%      \\
           & \tt force\_f     & (1.63) & (0.450) & (4.77\%)     \\
           & Fine-FFT     &   (1.41) & (0.389) & (4.13\%)     \\
           & Subtotal      & 1.06   & 0.292 & 3.09\%      \\
\hline
Total &                      & 34.2 & 9.43 & 100\%\\
\hline
\multicolumn{2}{l} {Optimal limit} & 21.7 & 6 & 63.6\%\\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
  \includegraphics[width=1.1\linewidth]{diff_power.pdf}
 \caption{power spectrum and cross correlation.}
\label{fig.ccc}
\end{figure}

 Cosmological simulation, PM method.
